---
title: "Thermal acclimation — methodology variants"
subtitle: "Every recipe against `original`, and `original` against the manuscript"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
    df-print: kable
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
knitr:
  opts_knit:
    root.dir: ".."
---

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(targets)

# Every table here is a lookup of what the pipeline wrote; nothing is refitted,
# so rendering is fast and cannot disagree with the run it describes.
#
# Outputs are reached through literal `tar_read()` calls: `tar_quarto()` scans
# this file for them and makes each a dependency, so the report re-renders when
# the results change rather than going quietly stale.
from_target <- function(path) {
  if (is.null(path) || !length(path) || !all(file.exists(path))) return(NULL)
  out <- read.csv(path, stringsAsFactors = FALSE)
  if (!nrow(out)) NULL else out
}
or_null <- function(expr) tryCatch(expr, error = function(e) NULL)

outcome  <- from_target(or_null(tar_read(variant_outcome_csv)))
settings <- from_target(or_null(tar_read(variant_settings_csv)))
siteyear <- from_target(or_null(tar_read(variant_siteyear_csv)))
skips    <- from_target(or_null(tar_read(variant_window_skips_csv)))
ts_qc    <- from_target(or_null(tar_read(ts_qc_csv)))
fill_cv  <- from_target(or_null(tar_read(fill_cv_csv)))
fill_sum <- from_target(or_null(tar_read(fill_summary_csv)))
feats    <- from_target(or_null(tar_read(growing_season_features_csv)))
recipes  <- from_target(or_null(tar_read(recipes_file)))

ORIG <- "data-proc-original"
read_if <- function(path) if (file.exists(path)) read.csv(path) else NULL
orig_total  <- read_if(file.path(ORIG, "outcome_temp.csv"))
orig_direct <- read_if(file.path(ORIG, "outcome_temp_water_gpp.csv"))

# Sampler noise floor: measured over identical repeated runs (reports/
# pipeline-report.qmd, and the replicate runs in ts-rework.html, F8). A TAS
# difference smaller than this means nothing.
TAS_NOISE <- 3e-4

skip_note <- function(what, why) {
  cat(sprintf("::: {.callout-warning}\n## %s not available\n\n%s\n:::\n\n", what, why))
}

# Recipes get their colour from their position in the registry, never from
# their rank in a particular table, so a recipe is the same colour in every
# figure and a filter that drops one never repaints the survivors. Five
# categorical slots from the validated reference palette, in its fixed order.
RECIPE_ORDER <- if (!is.null(recipes)) recipes$recipe_id else unique(settings$recipe_id)
PALETTE <- c("#2a78d6", "#eb6834", "#1baf7a", "#eda100", "#e87ba4", "#008300", "#4a3aa7", "#e34948")
recipe_cols <- setNames(PALETTE[seq_along(RECIPE_ORDER)], RECIPE_ORDER)
INK <- "#16181d"; INK2 <- "#4a5058"; GRID <- "#e2e5ea"

theme_report <- function() {
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = GRID, linewidth = 0.4),
      axis.title = element_text(colour = INK2),
      axis.text = element_text(colour = INK),
      strip.text = element_text(colour = INK, face = "bold", hjust = 0),
      legend.position = "top",
      legend.title = element_blank(),
      plot.title.position = "plot"
    )
}

fast_run <- !is.null(settings) && any(settings$fit_profile == "fast")
n_sites <- if (!is.null(settings)) length(unique(settings$site_ID)) else 0
```

```{r fast-callout}
#| echo: false
#| results: asis
if (fast_run) cat(
"::: {.callout-important}\n## This run used the `fast` fit profile\n\n",
"Some or all fits were made with fewer chains and iterations than the manuscript's ",
"configuration (`fit_settings(\"fast\")`). Every TAS below is a **smoke-test artefact** ",
"that shows the pipeline runs end to end; none of the numbers is a result. ",
"Re-run with `THERMAL_FIT=full` for anything you intend to read.\n:::\n\n")
if (n_sites > 0 && n_sites < 20) cat(
"::: {.callout-note}\n## Subset run\n\n",
sprintf("%d site(s) in this run. Site-level comparisons are meaningful; ", n_sites),
"cross-site summaries are not. `THERMAL_SITES=all` runs every site the pipeline handles.\n:::\n\n")
```

## Run metadata

```{r meta}
#| echo: false
git <- function(args) tryCatch(system2("git", args, stdout = TRUE), error = function(e) NA_character_)
tibble::tibble(
  item = c("Rendered", "Sites", "Recipes", "Models", "Fit profile(s)",
           "Git commit", "Git branch", "Working tree"),
  value = c(
    format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"),
    if (!is.null(settings)) paste(sort(unique(settings$site_ID)), collapse = ", ") else "none",
    if (!is.null(settings)) paste(unique(settings$recipe_id), collapse = ", ") else "none",
    if (!is.null(settings)) paste(unique(settings$model), collapse = ", ") else "none",
    if (!is.null(settings)) paste(unique(settings$fit_profile), collapse = ", ") else "none",
    paste(git(c("rev-parse", "--short", "HEAD")), collapse = ""),
    paste(git(c("rev-parse", "--abbrev-ref", "HEAD")), collapse = ""),
    if (length(git(c("status", "--porcelain")))) "modified" else "clean"
  )
)
```

## The recipes

A recipe is a named set of choices, one strategy per axis. The axes and what
each strategy does are documented in `docs/recipes.md`; the registry is
`data-core/recipes.csv`, and adding a variant is one row there.

```{r recipes}
#| echo: false
if (is.null(recipes)) skip_note("Recipe registry", "recipes_file target not built.") else
  recipes |> select(recipe_id, ts, season, bounds, swc, year_qc, description)
```

## `original` against the manuscript oracle

The `original` recipe is the manuscript's logic. Its TAS should match
`data-proc-original/` to within sampler noise; a larger difference is a real
difference between this code and the code the manuscript was made with.

```{r oracle}
#| echo: false
#| results: asis
orig_run <- if (!is.null(outcome)) filter(outcome, recipe_id == "original") else NULL
if (is.null(orig_run) || !nrow(orig_run)) {
  skip_note("Oracle comparison", "The `original` recipe is not in this run (THERMAL_RECIPES).")
} else if (is.null(orig_total)) {
  skip_note("Oracle comparison", "data-proc-original/outcome_temp.csv not found.")
} else {
  ref <- bind_rows(
    mutate(orig_total, model = "total"),
    if (!is.null(orig_direct)) mutate(orig_direct, model = "direct")
  ) |> select(site_ID, model, TAS_manuscript = TAS, TASp_manuscript = TASp)
  cmp <- orig_run |>
    select(site_ID, model, fit_profile, TAS, TASp) |>
    inner_join(ref, by = c("site_ID", "model")) |>
    mutate(
      delta = TAS - TAS_manuscript,
      delta_over_noise = round(abs(delta) / TAS_NOISE, 1),
      agrees = ifelse(abs(delta) <= TAS_NOISE, "yes", "**no**")
    ) |>
    arrange(model, site_ID)
  if (!nrow(cmp)) {
    skip_note("Oracle comparison", "No site in this run appears in the manuscript's outcome tables.")
  } else {
    cat(sprintf("%d of %d site-model pairs agree with the manuscript to within the sampler noise floor (%.0e).\n\n",
                sum(cmp$agrees == "yes"), nrow(cmp), TAS_NOISE))
    if (fast_run) cat("Under the `fast` profile disagreement is expected and says nothing.\n\n")
    print(knitr::kable(cmp, digits = 5))
  }
}
```

## TAS across recipes

For every site and model, the thermal-response strength under each recipe, the
change against `original`, and how many sampler-noise floors that change is.
A sign flip or a crossing of $p = 0.05$ is called out.

```{r tas-table}
#| echo: false
#| results: asis
if (is.null(outcome)) {
  skip_note("TAS comparison", "variant_outcome_csv not built.")
} else {
  base <- outcome |> filter(recipe_id == "original") |>
    select(site_ID, model, TAS_original = TAS, TASp_original = TASp)
  cmp_all <- outcome |>
    left_join(base, by = c("site_ID", "model")) |>
    mutate(
      delta = TAS - TAS_original,
      x_noise = round(abs(delta) / TAS_NOISE, 1),
      sign_flip = !is.na(TAS_original) & sign(TAS) != sign(TAS_original) & recipe_id != "original",
      sig_change = !is.na(TASp_original) & ((TASp < 0.05) != (TASp_original < 0.05)),
      note = paste0(ifelse(sign_flip, "sign flip; ", ""),
                    ifelse(sig_change, "p crosses 0.05; ", ""))
    )
  wide <- cmp_all |>
    select(site_ID, model, recipe_id, TAS) |>
    pivot_wider(names_from = recipe_id, values_from = TAS) |>
    arrange(model, site_ID)
  cat("### TAS by recipe (wide)\n\n")
  print(knitr::kable(wide, digits = 5))
  cat("\n\n### Change against `original`\n\n")
  cmp_all |>
    filter(recipe_id != "original") |>
    select(site_ID, model, recipe_id, TAS, TASp, TAS_original, delta, x_noise, note) |>
    arrange(model, site_ID, recipe_id) |>
    knitr::kable(digits = 5) |> print()
}
```

```{r tas-plot}
#| echo: false
#| fig-cap: "TAS by recipe. One point per recipe; the grey bar spans the recipes at each site. Recipes keep their colour across every figure."
if (!is.null(outcome) && nrow(outcome)) {
  d <- outcome |>
    mutate(recipe_id = factor(recipe_id, levels = RECIPE_ORDER),
           site_ID = factor(site_ID, levels = rev(sort(unique(site_ID)))))
  rng <- d |> summarise(lo = min(TAS), hi = max(TAS), .by = c(site_ID, model))
  ggplot(d, aes(y = site_ID)) +
    geom_segment(data = rng, aes(x = lo, xend = hi, yend = site_ID),
                 colour = GRID, linewidth = 2, lineend = "round") +
    geom_vline(xintercept = 0, colour = INK2, linewidth = 0.4) +
    geom_point(aes(x = TAS, colour = recipe_id), size = 3) +
    facet_wrap(~ model, scales = "free_x") +
    scale_colour_manual(values = recipe_cols, drop = FALSE) +
    labs(x = "TAS (slope of log respiration ratio on window soil temperature)", y = NULL) +
    theme_report()
}
```

## Provenance: what each recipe did at each site, and why

Every choice a recipe made, as the fit recorded it. `ts_reason` is the strategy's
own explanation; `ts_verdict`/`ts_flags` are the quality screen's; `fill_method`
and `fill_cv_rmse` are the blocked cross-validation's winner and its out-of-fold
error where a reconstruction was used. `ts_measured_synthetic` marks the sites
whose "measured" column is itself a reconstruction step 01 made (`estimate_Ts`,
`SITES_TS_FROM_TA_*`, `SITES_TS_SYNTHETIC`): the recipes select downstream of
those and cannot undo them, so at those sites every recipe but `noseason` fits
the same column. `fill_truth_synthetic` says the same of the CV's truth, and
`fill_degenerate` that the winner reproduced it to numerical zero.

```{r provenance}
#| echo: false
#| results: asis
if (is.null(settings)) skip_note("Provenance", "variant_settings_csv not built.") else {
  settings |>
    filter(model == first(model)) |>   # choices are model-independent except swc
    select(site_ID, recipe_id, ts_col, ts_reason, ts_verdict, ts_flags,
           any_of("ts_measured_synthetic"),
           fill_method, fill_cv_rmse, any_of(c("fill_degenerate", "fill_truth_synthetic")),
           bounds_strategy, tStart, tEnd,
           window_gStart, window_gEnd, control_year, fit_profile) |>
    arrange(site_ID, recipe_id) |>
    knitr::kable(digits = 3) |> print()
  cat("\n\nSoil water by model (the one model-dependent choice):\n\n")
  settings |> select(site_ID, recipe_id, model, swc_col, swc_reason) |>
    distinct() |> arrange(site_ID, recipe_id, model) |> knitr::kable() |> print()
}
```

## Structure: what each recipe fitted

Before any number, a recipe changes *which* (window × year) cells are fitted
and which year is the control. These are exact — the same loop the fits run.

```{r structure}
#| echo: false
#| results: asis
if (is.null(siteyear)) skip_note("Structure", "variant_siteyear_csv not built.") else {
  cells <- siteyear |>
    summarise(
      n_cells = sum(!is.na(alpha) | status %in% c("fitted", "not_fitted")),
      n_rejected = sum(!status %in% c("fitted", "not_fitted")),
      .by = c(site_ID, recipe_id, model)
    )
  ctrl <- settings |> select(site_ID, recipe_id, model, control_year, nwindow)
  ctrl_orig <- ctrl |> filter(recipe_id == "original") |>
    select(site_ID, model, control_year_original = control_year)
  cells |>
    left_join(ctrl, by = c("site_ID", "recipe_id", "model")) |>
    left_join(ctrl_orig, by = c("site_ID", "model")) |>
    mutate(control_changed = ifelse(control_year != control_year_original, "**yes**", "")) |>
    arrange(site_ID, model, recipe_id) |>
    knitr::kable() |> print()
  if (!is.null(skips) && nrow(skips)) {
    cat("\n\nWindow skips by reason:\n\n")
    skips |> count(recipe_id, model, reason) |> pivot_wider(names_from = reason, values_from = n, values_fill = 0) |>
      knitr::kable() |> print()
  }
}
```

## Is season detection redundant? The `noseason` test

The hypothesis: push the whole year through and the fit-stage guards (soil
temperature below 2 °C, too few observations, non-positive respiration) will
discard the same windows season detection would have. If that were true, the
cells `noseason` fits would lie almost entirely inside the detected season.

```{r noseason}
#| echo: false
#| results: asis
ns <- if (!is.null(siteyear)) filter(siteyear, recipe_id == "noseason") else NULL
if (is.null(ns) || !nrow(ns)) {
  skip_note("noseason test", "The `noseason` recipe is not in this run.")
} else if (is.null(feats)) {
  skip_note("noseason test", "growing_season_features_csv not built.")
} else {
  w <- ns |>
    filter(status %in% c("fitted", "not_fitted")) |>
    mutate(w_start = as.numeric(sub("_.*", "", window)),
           w_end = as.numeric(sub(".*_", "", window))) |>
    left_join(select(feats, site_ID, gStart, gEnd), by = "site_ID") |>
    mutate(where = case_when(
      w_end < gStart | w_start > gEnd ~ "entirely outside the detected season",
      w_start >= gStart & w_end <= gEnd ~ "entirely inside",
      TRUE ~ "straddles a boundary"
    ))
  tab <- w |> count(site_ID, model, where) |>
    group_by(site_ID, model) |> mutate(share = round(n / sum(n), 3)) |> ungroup()
  print(knitr::kable(tab))
  outside <- tab |> filter(where == "entirely outside the detected season")
  if (nrow(outside)) cat(sprintf(
    "\n\nAcross this run, **%.0f%%** of the cells `noseason` fitted lie entirely outside the detected growing season. ",
    100 * sum(outside$n) / sum(tab$n)),
    "If the hypothesis held, this would be close to zero. The out-of-season cells that survive are those the guards do not remove: ",
    "shoulder-season windows with soil above 2 °C and positive nighttime respiration.\n\n")
  cat("\n\nWhat this test does *not* change: the detected season still sets the year gap scan and the control year. ",
      "A fully season-free variant needs a replacement rule for both; see `docs/recipes.md`.\n\n")
}
```

## The soil-temperature reconstruction each site would use

Blocked cross-validation (leave-one-year-out) of every candidate method at each
site. The winner is the method a `memory_fill` recipe attaches as `TS_memfill`
wherever the quality screen fails the measured column. `lm_ta_pos` is the
pipeline's pre-existing `TS ~ TA` default, scored on the same folds.

```{r fill}
#| echo: false
#| results: asis
if (is.null(fill_sum)) skip_note("Fill", "fill_summary_csv not built.") else {
  cat("### Decision per site\n\n")
  print(knitr::kable(fill_sum, digits = 3))
  if (!is.null(fill_cv)) {
    cat("\n\n### Cross-validation table\n\n")
    fill_cv |>
      select(site_ID, method, rmse, cell_mean_rmse, within_sd_ratio,
             across_year_spread_ratio, amp_inflation, lag_pred_h) |>
      arrange(site_ID, rmse) |>
      knitr::kable(digits = 3) |> print()
  }
}
```

```{r fill-plot}
#| echo: false
#| fig-cap: "Out-of-fold RMSE by method and site. Lower is better; the incumbent `lm_ta_pos` is marked."
if (!is.null(fill_cv) && nrow(fill_cv)) {
  d <- fill_cv |> mutate(incumbent = method == "lm_ta_pos")
  ggplot(d, aes(x = rmse, y = reorder(method, -rmse))) +
    geom_col(aes(fill = incumbent), width = 0.6) +
    facet_wrap(~ site_ID) +
    scale_fill_manual(values = c(`FALSE` = "#2a78d6", `TRUE` = "#eb6834"),
                      labels = c(`FALSE` = "candidate", `TRUE` = "incumbent (TS ~ TA)")) +
    labs(x = "out-of-fold RMSE (°C), leave-one-year-out", y = NULL) +
    theme_report()
}
```

## Soil-temperature quality screen

The verdict the `screen_best` and `memory_fill` strategies branch on, per site,
with the diagnostics behind it. `amp_ratio` is the diurnal amplitude of soil
temperature over that of air (buried soil damps; near 1 is not soil);
`lag_unwrapped` is the phase lag in hours (soil lags air; zero is not soil).

```{r tsqc}
#| echo: false
#| results: asis
if (is.null(ts_qc)) skip_note("TS quality", "ts_qc_csv not built.") else {
  ts_qc |> select(site_ID, verdict, flags, coverage, amp_ratio, lag_unwrapped,
                  stuck_frac, mean_offset, n) |>
    knitr::kable(digits = 3) |> print()
}
```

## How to read this

- `original` is the manuscript. Everything else is a *variant* and is judged
  against it, and `original` is judged against `data-proc-original/`.
- A TAS difference below the noise floor (`r format(TAS_NOISE)`) is not a
  difference. The replicate runs in `ts-rework.html` (F8) measured this floor
  at between $10^{-5}$ and $6 \times 10^{-4}$.
- A recipe's number is only as trustworthy as the choice behind it, which is
  why the provenance table sits next to the results rather than in an appendix.
- Under the `fast` profile nothing here is a result.
